---
title: "midterm_key_oct_8"
author: "Rob Wells"
date: "2025-10-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam - Fall 2025

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

# Load tidyverse
```{r}
library(tidyverse)
```

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade.

## Task 1: Read the data**
```{r}
midterm_set <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```


## Task 2: Calculate the mean and standard deviation maximum and minimum of compliance scores. Put the results in a new dataframe called "compliance_stats." 
```{r}

compliance_stats <- midterm_set %>%
summarise(
mean_compliance = mean(compliance_score, na.rm = TRUE),
sd_compliance = sd(compliance_score, na.rm = TRUE),
max_compliance = max(compliance_score, na.rm = TRUE),
min_compliance = min(compliance_score, na.rm = TRUE)
)
```

## Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a headline. Include a headline and label the x and y axes.
```{r}
ggplot(midterm_set, aes(x = compliance_score)) +
geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
geom_vline(aes(xintercept = mean(compliance_score, na.rm = TRUE)),
color = "red", linetype = "dashed", linewidth = 1) +
labs(
title = "Histogram of Compliance Scores",
x = "Compliance Score",
y = "Count"
) +
theme_minimal()

```

## Task 4: Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? 

**Write your answer here

ANSWER: This dataset has scores clustered closely around the mean of about 96, with a standard deviation of roughly 5.8. This reveals that the health inspections of the restaurants are doing quite great, with only a few falling much below the average. The point of interest is that despite the fact that restaurants in the county usually have high standards, the low-scoring places might be the ones to give rise to inquiries regarding the reasons behind their bad results.


## Part 2: High School Athletics (30 points)

The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and gender. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv`

# Task 5: Load the high school participation data
```{r}
hs_set <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")
```


#Task 6: Add two columns called total and girls_pct.  
Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. 
```{r}
hs_set <- hs_set %>%
mutate(
total = boys + girls,
girls_pct = (girls / total) * 100
)


```

#Task 7: Calculate correlation between boys and girls participation
```{r}
cor(hs_set$boys, hs_set$girls)
```

#Task 8: Create a scatterplot showing this relationship, adding a line of best fit. Include a headline and label the x and y axes.
```{r}
hs_set |>
ggplot() +
geom_point(aes(x = boys, y = girls)) +
geom_smooth(aes(x = boys, y = girls), method = "lm") +
expand_limits(x = 0, y = 0) +
labs(
title = "Relationship Between Boys and Girls High School Sports Participation in Maryland (2024)",
x = "Boys Participation",
y = "Girls Participation"
) +
theme_minimal()

```


# Task 9: In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about gender equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

**Write your answer here:**
The scatterplot shows a strong positive correlation indicating that schools with a higher number of boys participating also tend to have a higher number of girls participating. This implies a more or less equal gender distribution over most of the Maryland districts. Districts that are situated below the line have girls participating less than boys, whereas districts above the line have girls participating more then males.

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:

#Task 10: Load the car thefts data
```{r}
carthefts_set <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

#Task 11: Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate 
```{r}

carthefts_set <- carthefts_set %>%
mutate(
car_theft_rate_2023 = (`2023` / population) * 1000
)

View(carthefts_set)

```

#Task 12: Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median? 
```{r}
median_rate <- median(carthefts_set$car_theft_rate_2023, na.rm = TRUE)
total_thefts <- sum(carthefts_set$`2023`, na.rm = TRUE)
above_median <- carthefts_set %>%
  filter(car_theft_rate_2023 > median_rate)

median_rate
total_thefts
view(above_median)

```
ANSWER: There are 12 counties that have rates above the median. They are Anne Arundel County, Anne Arundel County, Baltimore County, Cecil County, Charles County, Dorchester County, Howard County, Montgomery County, Prince George's County, St. Mary's County, Washington County, and Wicomico County. All 12 of these counties have rates higher then the 0.9580638 median rate. 

#Task 13: Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? 

**Write your answer here:**
ANSWER: In Maryland, car thefts are not evenly spread, as 12 counties have rates that are above the stateside median which is about 0.96 thefts per 1,000 residents. Urban and suburban areas such as Prince George’s, Baltimore, and Montgomery counties are among the areas with above-average rates of car thefts which suggest that densely populated or developed areas are more likely to encounter difficulties with vehicle theft. On the other hand, few rural counties have very low reports of car thefts, meaning that there is a great geographical variation in the risk of car theft.

Lede: Car thefts in Maryland are hitting some counties much harder than others, with 12 counties reporting rates above the state median.

## Part 4: Data Analysis Scenario (20 points)

#Task 14: Essay Question.

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)
2. What visualizations would help readers understand the trends? (5 points)
3. What additional context or data would you need to make this a complete story? (5 points)

**Write your answer here:**
ANSWERS: 
1. To verify the claim, I would calculate the mean, median, and standard deviation of monthly response times for police, fire, and ambulance services over the past year, and compare them to the previous year's data.
2. Some visualizations that would help readers understand the trends would be histograms and scatterplots to make the data easier to comprehend. Graphs are a great way to illustrate to viewers how they correlate with different variables in the situation rather then just looking at different numerical values.
3. Some additional context that would make this a complete story could be the amount of fire responders working for the police, fire department, and ambulances. This would be important because if there has been a steady decline of workers then that could be a reason why the response times have gotten significantly worse over the past year. Other important factors could be how many hours a week these first responders work or even if there pay has declined over the years. 

# Excellent Response


### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Submit both your code and written responses
- Load this test to GitHub and submit the link in Elms

Good luck!
