---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)



```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities %>%
  mutate(
    violent_rate_2019  = (violent2019  / pop2019) * 1000,
    property_rate_2019 = (property2019 / pop2019) * 1000,
    violent_rate_2020  = (violent2020  / pop2020) * 1000,
    property_rate_2020 = (property2020 / pop2020) * 1000
  )

View(md_cities_rates)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Baltimore had the highest violent rate in 2019. Ocean city was second while Williamsport was third.Regarding the lowest, it was it was a bunch of cities tied with having 0 violent rate including some such as University Park, Hebon, and Oxford.


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates %>%
  mutate(
    violent_pct_change  = ((violent_rate_2020 - violent_rate_2019) / violent_rate_2019) * 100,
    property_pct_change = ((property_rate_2020 - property_rate_2019) / property_rate_2019) * 100
  )

View(md_cities_change)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

While the general tendency for many cities is to show modest shifts in violence or property crime rates between the years of 2019 and 2020, there are a handful of jurisdictions with outsized increases or decreases. Baltimore City, for example, saw a 12.9% decline in violent crime and a further decline of 25.9% in property crime, whereas smaller towns such as Brookeville, Chevy Chase Sector V, and Oxford recorded infinitely large or undefined percentage changes-one-way increases or decreases because the crime incidences were minutely negligible in one of the years.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Baltimore City leads every single Maryland city in property crime and violent crime in 2019 and 2020. 

The year 2020 saw an extension of the trend of crime in Baltimore, which puts in place the prerogative of highest rates of violent and property crimes in the Maryland cities. While large cities like Baltimore show meaningful trends in crime rates, smaller towns can experience dramatic percentage swings due to their low populations, which can make changes look extreme. It is an important data point because it highlights where crime occurs in highest concentration, thus guiding policymakers, law enforcement, and residents in focusing their resources and considering public safety challenges across the state.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.


### Task 10: Your task will be to take the violent crime rates for 2020 and do three things:


```{r}
md_cities_percentiles <- md_cities_rates 
  mutate(percentile_2020 = ntile(violent_rate_2020, 100))
cities_above_90th <- md_cities_percentiles %>%
  filter(percentile_2020 > 90) %>%
  nrow()

cities_above_90th

# Part 2:

iqr_violent_2020 <- md_cities_rates %>%
  summarize(iqr_val = IQR(violent_rate_2020, na.rm = TRUE))

iqr_violent_2020
```
