---
title: "HW7_MandMs"
date: 10/1/2025
---


In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load an M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
 dataset <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10. Be sure to use set.seed(1234) before you start.
```{r}
set.seed(1234)
sample5 <- dataset %>% sample_n(5)
sample10 <- dataset %>% sample_n(10)
sample5
sample10
```


### Task 4: Calculate the mean, median and standard deviation of red M&Ms from each of the samples.
```{r}
sample5 %>% summarise(mean_red = mean(red), median_red = median(red), sd_red = sd(red))
sample10 %>% summarise(mean_red = mean(red), median_red = median(red), sd_red = sd(red))
```


### Task 5: Calculate the mean, median and standard deviation of red M&Ms from the population

```{r}
dataset %>% summarise(mean_red = mean(red), median_red = median(red), sd_red = sd(red))
```

### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Include a headline. Use a consistent binwidth.

```{r}
ggplot(sample5, aes(x = red)) +
  geom_histogram(binwidth = 5, fill = "red", color = "black") +
  ggtitle("Distribution of Red M&Ms (Sample of 5)")

ggplot(sample10, aes(x = red)) +
  geom_histogram(binwidth = 5, fill = "red", color = "black") +
  ggtitle("Distribution of Red M&Ms (Sample of 10)")

ggplot(dataset, aes(x = red)) +
  geom_histogram(binwidth = 5, fill = "red", color = "black") +
  ggtitle("Distribution of Red M&Ms (Full Dataset)")

```


### Task 7: Answer questions


Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?

What does this exercise tell you about the relationship between sample size and estimate accuracy?

How does random sampling introduce variation into statistical estimates?

If you repeated this process, would the samples always be identical? Why or why not?


The means and standard deviations from the bigger sample (10) are closer to population values than the ones from the smaller sample (5). The histograms indeed prove the point that as sample size grows, the distribution starts to look more and more like that of the population. Hence, agreed, a larger sample size gives a better picture of the overall pattern in the data.

This exercise shows that estimates from larger samples tend to be reliable and closer to true population values. Small samples, on the other hand, can vary to a great extent, but larger samples smooth out these random fluctuations.

There is variation introduced by random sampling since each sample is a different combination of observed units from the population. That variation in turn causes the sample statistics—mean and standard deviation—to differ slightly every time.

These samples would not be exactly the same with repeated trials because the sampling is random. There would be differences in the specific numbered rows chosen each time and just a little variation in the observed results, but with larger samples, these differences would shrink and estimates would hold close to population values.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
