---
title: "final project"
author: "Justin Corman" 
output: html_document
date: "2025-11-12"
---
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
```
```{r}
# Load + clean data
billboard <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQVR7XE8UGPh9EYKpSLzrNA0kXqcQo3AONREuc6mAAaHAu-8ia01STi_HrHgATfgJ5n9xUAUHPSl9ou/pub?gid=1077686609&single=true&output=csv") |> 
  clean_names() |> 
  mutate(date_new = dmy(date)) |> 
  mutate(
    date_new = if_else(year(date_new) > 2024, date_new - years(100), date_new),
    year = year(date_new)
  ) |> 
  select(date_new, year, everything())
```
```{r}
# Summary statistics
music_stats <- billboard %>%
  summarise(
    mean_bpm = mean(bpm, na.rm = TRUE),
    sd_bpm = sd(bpm, na.rm = TRUE),
    min_bpm = min(bpm, na.rm = TRUE),
    max_bpm = max(bpm, na.rm = TRUE),
    median_bpm = median(bpm, na.rm = TRUE)
  )

music_stats

# Create eras + decades
billboard <- billboard %>%
  mutate(
    era = case_when(
      year >= 1970 & year <= 1999 ~ "1970_1999",
      year >= 2010                ~ "2010_2020s",
      TRUE                        ~ "other"
    ),
    decade = case_when(
      year >= 1970 & year <= 1979 ~ "1970s",
      year >= 1980 & year <= 1989 ~ "1980s",
      year >= 1990 & year <= 1999 ~ "1990s",
      year >= 2010 & year <= 2019 ~ "2010s",
      year >= 2020                ~ "2020s",
      TRUE                        ~ NA_character_
    )
  )
```
```{r}
# T-test prep
bpm_eras <- billboard %>%
  filter(era %in% c("1970_1999", "2010_2020s")) %>%
  drop_na(bpm)

bpm_summary_by_era <- bpm_eras %>%
  group_by(era) %>%
  summarise(
    mean_bpm = mean(bpm),
    sd_bpm   = sd(bpm),
    n        = n(),
    .groups  = "drop"
  )

bpm_1970_1999  <- bpm_eras %>% filter(era == "1970_1999")  %>% pull(bpm)
bpm_2010_2020s <- bpm_eras %>% filter(era == "2010_2020s") %>% pull(bpm)

# T-test (modern ≥ 10 BPM faster)
t_bpm <- t.test(
  bpm_2010_2020s,
  bpm_1970_1999,
  alternative = "greater",
  mu = 10
)

t_bpm
```
```{r}
# ANOVA across decades
billboard_decades <- billboard %>%
  filter(!is.na(decade), year >= 1970) %>%
  drop_na(bpm)

anova_model <- aov(bpm ~ decade, data = billboard_decades)
summary(anova_model)

tukey_bpm <- TukeyHSD(anova_model)
tukey_bpm

# Visualization
ggplot(billboard_decades, aes(x = decade, y = bpm)) +
  geom_boxplot(outlier.alpha = 0.4) +
  geom_jitter(width = 0.15, alpha = 0.2) +
  labs(
    title = "BPM of Billboard Hot 100 #1 Songs by Decade",
    x = "Decade",
    y = "Beats Per Minute (BPM)"
  ) +
  theme_minimal()
```


Statistics for Music BPM
mean:	115.8562
SD: 24.55999 
min:16
max:206
median:115


PART 2: 
I focused on whether modern #1 hits are meaningfully faster than older ones. I compared BPM values from two eras: 1970–1999 and 2010–2020s. My hypotheses were:

UPDATED HYPOTHESIS'S:
Ho: The average BPM of 2010–2020s #1 songs is at most 10 BPM higher than the average BPM from 1970–1999.

H1: The average BPM of 2010–2020s #1 songs is more than 10 BPM higher than the 1970–1999 average.
These hypotheses fit a one-sided two-sample t-test, which compares mean BPM across two independent time periods.

T-TEST:

The t-test compares whether modern #1 hits have increased in BPM by more than 10 BPM compared to older hits. The summary statistics show that recent songs have a higher mean BPM, but the t-test evaluates whether that difference is large enough to meet the 10-BPM threshold.

If the p-value < 0.05, I reject the null hypothesis and conclude that modern #1 hits are significantly more than 10 BPM faster.

If the p-value > 0.05, I fail to reject the null hypothesis, meaning the increase is not statistically greater than 10 BPM.
After reviewing my output, the p-value was 0.9853, so I did not reject the null hypothesis.

ANOVA TEST: 

I also ran a one-way ANOVA to test whether BPM differs across decades (1970s–2020s). The ANOVA checks whether at least one decade has a significantly different mean tempo. A small p-value would indicate that tempo changes across decades are statistically meaningful. The Tukey test identifies which decades differ from one another.
Based on my results, the ANOVA p-value was 0.0197, which means BPM values across decades do show statistically significant differences.

BOXPLOT: 

The boxplot of BPM by decade helps visualize how song tempo has shifted over time. Higher median lines for the 2010s or 2020s would visually support the idea that modern hits tend to be faster. The overlap and spread also show that while the overall trend changes, there is still wide variation in tempo within each decade.



PART 1: 

--What does it contain and what time period does it cover?

ANSWER HERE: The Billboard Hot 100 Number Ones dataset contains detailed information about every song that has reached the #1 position on the Billboard Hot 100 chart. Each row represents a unique chart-topping song and includes variables such as the title, artist, release year, and BPM just to name a few. The dataset covers hit songs from 1958 through the early 2020s, providing more than sixty years of musical trends and allowing for a comprehensive look at how #1 hits have evolved over time.

--Which columns will be most important for your analysis?

ANSWER HERE:The columns that will matter most for my analysis are those that directly describe the musical qualities of each song, because they help reveal patterns in what makes a song successful. In particular, BPM, Energy, and Danceability are likely to be the key variables, as each one captures an aspect of a song’s sound or structure. The Year column is also important since it allows for comparisons across decades.

--What's missing that you might need or want?

ANSWER HERE: Several valuable pieces of information are missing from the dataset that could help explain why certain songs make it to #1. For example, there is no data on streaming counts, radio airplay, or even TikTok engagement that could show why some are heavily influenced on the modern chart performances. Having these variables would make it easier to understand the cultural forces behind a hit, but they are not included in the dataset, so the analysis must rely only on musical and structural characteristics.

Proposed hypothesis: 

Null Hypothesis (Ho):
The average BPM of #1 songs in the 2010s–2020s is less than or equal to the average BPM of #1 songs from 1970–1999.

Alternative Hypothesis (H1):
The average BPM of #1 songs in the 2010s–2020s is at least 10 BPM higher than the average BPM of #1 songs from 1970–1999.

Questions I have: 

ANSWER HERE: Some questions that I still have about the dataset include whether musical characteristics like BPM or Energy truly changed in a meaningful way across decades, or whether the rise of social media has altered what types of songs reach #1. I also want to explore whether shorter songs are becoming more common at the top of the charts and whether certain genres have dominated specific time periods. These questions can guide the next steps of my analysis as I learn to use visualization and statistical tools to confirm or reject potential story ideas within the dataset.
