---
title: "Regression and Beers"
author: "Wells"
date: "2025-12-02"
output: html_document
---

This is another exercise to break down how a simple linear regression works and how to use it in real life.
The topic: beer.

This following exercise comes from Diez, D., Cetinkaya-Rundel, M., & Barr, C. (2024). OpenIntro Statistics (4th ed.). http://www.openintro.org/redirect.php?go=os&referrer=os4_pdf, p 335


Many people believe that gender, weight, drinking habits, and many other factors are much more important in predicting blood alcohol content (BAC) than simply considering the number of drinks a person consumed. Here we examine data from sixteen student volunteers at Ohio
State University who each drank a randomly assigned number of cans of beer. These students were evenly divided between men and women, and they differed in weight and drinking habits. Thirty minutes later, a police officer measured their blood alcohol content (BAC) in grams of alcohol per deciliter of blood.



Load libraries
```{r}
library(tidyverse)
```


Load data
```{r}
bac_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/activities/bac%20regression/bac_data.csv")

bac_data

#Student #1 drank 5 beers and has a blood alcohol level of 0.13, etc
```


# Linear Regression
```{r}
# Run the regression
model <- lm(bac ~ beers, data = bac_data)

# Display regression results
cat("\n\nRegression Results:\n")
print(summary(model))
```

 Plot the data and the model line
```{r}
# Create regression equation text
eq_text <- sprintf("BAC = %.4f + %.4f * beers", 
                   coef(model)[1], coef(model)[2])


ggplot(bac_data, aes(x = beers, y = bac)) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1) +
  labs(
    x = "Number of Beers",
    y = "Blood Alcohol Content (BAC)",
    title = "BAC vs Number of Beers Consumed"
  ) +
  annotate("text", x = min(bac_data$beers), y = max(bac_data$bac), 
           label = eq_text, hjust = 0, vjust = 1, size = 4) +
  theme_minimal()

```

**Task 1: Interpret the results of this regression. Answer the following questions:**

a) What the predictor or independent variable? What is the response or dependent variable? 
b) What is the regression model asking and what does it tell us?
c) Do the data provide strong evidence that drinking more cans of beer is associated with an increase in blood alcohol? 
d) State the null and alternative hypotheses, report the p-value, and state your conclusion.


**Answer:**
a. The independent variable is the Number of beers consumed abd tge dependent variable is the blood alcohol content
b. The model asks whether BAC can be predicted from the number of beers consumed and if drinking more beers tends to increase BAC. The model tells us that the Intercept = 0.0060: Predicted BAC when beers = 0. That the slope = 0.0162: For each additional beer consumed, BAC increases by 0.0162 on average. So the model shoes a positive linear relationship: more beers → higher BAC.
c. Based on the scatterplot and upward-sloping regression line, there appears to be a positive association and it shows strong evidence. The estimated slope is 0.01618, meaning each additional beer is associated with an increase of 0.01618 BAC units.The p-value for the slope is 0.000524, which is far below any standard significance level (0.05, 0.01, even 0.001).
d. The null hypothesis states that the slope between number of beers and BAC is zero, meaning there is no relationship, while the alternative hypothesis states that the slope is greater than zero, meaning BAC increases as the number of beers increases. Based on the regression output, the p-value for the slope is 0.000524, which is far smaller than 0.05. Therefore, we reject the null hypothesis and conclude that there is strong statistical evidence that drinking more beers is associated with an increase in blood alcohol content.



